---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# EvoPhylo: Pre- and Postprocessing of Morphological Data from Relaxed Clock Bayesian Phylogenetics

A package to perform automated morphological character partitioning for phylogenetic analyses and analyze macroevolutionary parameter outputs from clock (time-calibrated) Bayesian inference analyses. `EvoPhylo` was initially released for pre- and postprocess data using the software [Mr. Bayes](https://nbisweden.github.io/MrBayes/), but since version 0.3 it also handles data pre- and post-processing for the software package [BEAST2](http://www.beast2.org/).

The ideas and rationale behind the original functionality and objectives of the analyses available in this package were first presented by [Simões & Pierce (2021)](https://doi.org/10.1038/s41559-021-01532-x). It's current functionality is described in detail by [Simões, Grifer, Barido-Sottani & Pierce (2022)](https://doi.org/10.32942/osf.io/3mkf9).

## Installing package **EvoPhylo**

Install the latest release version (v. 0.3) directly from CRAN:

```{r, eval = FALSE}
install.packages("EvoPhylo")
```

or development version from Github:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("tiago-simoes/EvoPhylo")
```

## Tutorials

See `vignette("char-part")`,`vignette("data_treatment")`, `vignette("rates-selection")`, `vignette("fbd-params")`, and `vignette("offset_handling")`, for step-by-step guides on using `EvoPhylo` to perform these analyses, also available on the [EvoPhylo website](https://tiago-simoes.github.io/EvoPhylo/).


## Authors

## Authors (including current developers)

*   [**Tiago R. Simões**](https://tiago-simoes.com/): *Conceptualization and code development* - [Github Profile](https://github.com/tiago-simoes)
*   [**Noah Greifer**](https://github.com/ngreifer): *Code development*
*   [**Joëlle Barido-Sottani**](https://github.com/bjoelle): *Code development*
*   [**Stephanie E.Pierce**](https://projects.iq.harvard.edu/spierce/home): *Project supervision*

## License

This project is licensed under General Public License, version 2.
