---
title: "Evolutionary Rates & Selection Mode"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
vignette: |
  %\VignetteIndexEntry{Evolutionary Rates & Selection Mode} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

This vignette explains how to extract evolutionary rate parameters estimated from relaxed clock Bayesian inference analyses produced by the program [Mr. Bayes](https://nbisweden.github.io/MrBayes/). It also shows how to use evolutionary rate based inference of selection mode (strength) adapted to clock-based rates, as introduced by @simões2021.

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

## Evolutionary Rates Statistics and Plots


In this section, we will extract evolutionary rate parameters from each node from a Bayesian clock (time-calibrate) summary tree produced by [Mr. Bayes](https://nbisweden.github.io/MrBayes/). The functions below will store them in a data frame, produce summary statistics tables, and create different plots showing how rates are distributed across morphological partitions and clades.


Load the **EvoPhylo** package

```{r}
library(EvoPhylo)
```

### 1. Get rates from the clock tree and create a rate table 

First, import a Bayesian clock tree using `treeio`'s function `read.mrbayes()` (= `read.beast()`).

```{r, eval = FALSE}
## Import summary tree with three clock partitions produced by 
## Mr. Bayes (.t or .tre files) from your local directory
tree3p <- treeio::read.mrbayes("Tree3p.t")
```

Below, we use the example tree `tree3p` that accompanies `EvoPhylo`.

```{r}
data(tree3p)
```

Subsequently, using `get_clockrate_table()`, users can extract mean or median rate values for each node in the summary tree that were annotated by [Mr. Bayes](https://nbisweden.github.io/MrBayes/) when creating the summary tree with [Mr. Bayes](https://nbisweden.github.io/MrBayes/) "sumt" command. These mean or median rate values are calculated by [Mr. Bayes](https://nbisweden.github.io/MrBayes/) taking into account all trees from the posterior sample. This works for any summary tree produced by [Mr. Bayes](https://nbisweden.github.io/MrBayes/): a majority rule consensus or the fully resolved maximum compatible tree (the latter is used in the examples here).

Please note that analyses must have reached the stationarity phase and independent runs converging for the summary statistics in each node to be meaningful summaries of the posterior sample.

```{r}
## Get table of clock rates with summary stats for each node in 
## the tree for each relaxed clock partition 
rate_table_means_no_clades3 <- get_clockrate_table(tree3p, summary = "mean")
```


### 2. Export the rate table 

This is a necessary step to subsequently open the rate table spreadsheet locally (e.g., using Microsoft Office Excel) and customize the table with clade names associated with with each node in the tree for downstream analysis.

```{r, eval = FALSE}
## Export the rate tables
write.csv(rate_table_means_no_clades3, file = "RateTable_Means3.csv")
```


### 3. Plot tree node labels

To visualize the node values in the tree, you can use `ggtree()`.

```{r, fig.width=8, fig.height=8, fig.align = "center", out.width = "70%", message=FALSE,warning=FALSE}
## Plot tree node labels
library(ggtree)
tree_nodes <- ggtree(tree3p, branch.length = "none", size = 0.05) +
  geom_tiplab(size = 2, linesize = 0.01, color = "black", offset = 0.5) +
  geom_label(aes(label = node), size = 2, color="purple", position = "dodge")
tree_nodes
```
```{r, eval = FALSE}
## Save your plot to your working directory as a PDF
ggplot2::ggsave("Tree_nodes.pdf", width = 10, height = 10)
```


### 4. Get summary statistics table and plots 

Import the rate table with clade membership (new "clade" column added)

```{r,eval=FALSE}
## Import rate table with clade membership (new "clade" column added) 
## from your local directory
rate_table_clades_means3 <- read.csv("RateTable_Means3_Clades.csv", header = TRUE)
```

Below, we use the rate table with clade membership `rate_table_clades_means3` that accompanies `EvoPhylo`.

```{r}
data(rate_table_clades_means3)

head(rate_table_clades_means3)
```

Obtain summary statistics table and plots for each clade by clock using `clockrate_summary()`. Supplying a file path to `file` save the output to that file.

```{r, eval = FALSE}
## Get summary statistics table for each clade by clock 
clockrate_summary(rate_table_clades_means3,
                  file = "Sum_RateTable_Means3.csv")
```

```{r, echo = FALSE}
t1 <- clockrate_summary(rate_table_clades_means3, digits = 2) 
kableExtra::kbl(t1, caption = "Rate table summary statistics") |> 
  kableExtra::kable_styling(font_size = 15, full_width = FALSE,
                            bootstrap_options = "striped", "condensed")
```


### 5. Plot rates by clock partition and clade

Plot distributions of rates by clock partition and clade with `clockrate_dens_plot()`.

```{r, fig.width=8, fig.height=5, fig.align = "center", out.width = "100%"}
## Overlapping plots
clockrate_dens_plot(rate_table_clades_means3, stack = FALSE,
                    nrow = 1, scales = "fixed")
```

Sometimes using stacked plots provides a better visualization as it avoids overlapping distributions.

```{r, fig.width=8, fig.height=5, fig.align = "center", out.width = "100%"}
## Stacked plots
clockrate_dens_plot(rate_table_clades_means3, stack = TRUE,
                    nrow = 1, scales = "fixed")
```

It is also possible to append extra layers using `ggplot2` function, such as for changing the color scale. Below, we change the color scale to be the Viridis scale.

```{r, fig.width=8, fig.height=5, fig.align = "center", out.width = "100%"}
## Stacked plots with viridis color scale
clockrate_dens_plot(rate_table_clades_means3, stack = TRUE,
                    nrow = 1, scales = "fixed") +
  ggplot2::scale_color_viridis_d() +
  ggplot2::scale_fill_viridis_d()
```

### 6. Rate linear models

We can also plot linear model regressions between rates from two or more clocks with `clockrate_reg_plot()`.

```{r, fig.width=8, fig.height=8, fig.align = "center", out.width = "70%"}
## Plot regressions of rates from two clocks
p12 <- clockrate_reg_plot(rate_table_clades_means3, clock_x = 1, clock_y = 2)
p13 <- clockrate_reg_plot(rate_table_clades_means3, clock_x = 1, clock_y = 3)
p23 <- clockrate_reg_plot(rate_table_clades_means3, clock_x = 2, clock_y = 3)

library(patchwork) #for combining plots
p12 + p13 + p23 + plot_layout(ncol = 2)
```
```{r, eval = FALSE}
## Save your plot to your working directory as a PDF
ggplot2::ggsave("Plot_regs.pdf", width = 8, height = 8)
```

### Rates from single clock analysis

You can also explore clock rates for summary trees including a single clock shared among all character partitions (or an unpartitioned analysis):

```{r,eval = FALSE}
## Import summary tree with a single clock partitions produced by 
## Mr. Bayes (.t or .tre files) from examples directory
tree1p <- treeio::read.mrbayes("Tree1p.t")
```

Below, we use the example tree `tree1p` that accompanies `EvoPhylo`.

```{r}
data(tree1p)
```

Then, get table of clock rates with summary stats for each node in the tree for each relaxed clock partition.

```{r}
rate_table_means_no_clades1 <- get_clockrate_table(tree1p, summary = "mean")
```

```{r, eval = FALSE}
## Export the rate tables
write.csv(rate_table_means_no_clades1, file = "RateTable_Means1.csv")

## Import rate table after adding clade membership (new "clade" column added)
rate_table_clades_means1 <- read.csv("RateTable_Means1_Clades.csv", header = TRUE)
```
```{r}
data(rate_table_clades_means1)
```
```{r,eval = FALSE}
## Get summary statistics table for each clade by clock 
clockrate_summary(rate_table_clades_means1,
                  file = "Sum_RateTable_Medians1.csv")
```
```{r, echo = FALSE}
t1 <- clockrate_summary(rate_table_clades_means1, digits = 2) 
kableExtra::kbl(t1, caption = "Rate table summary statistics") |> 
  kableExtra::kable_styling(font_size = 15, full_width = FALSE,
                            bootstrap_options = "striped", "condensed")
```

```{r, fig.width=8, fig.height=5, fig.align = "center", out.width = "100%"}
## Stacked plots with viridis color scale
clockrate_dens_plot(rate_table_clades_means1, stack = TRUE,
                    nrow = 1, scales = "fixed") +
  ggplot2::scale_color_viridis_d() +
  ggplot2::scale_fill_viridis_d()
```

---

## Selection Mode

In this section, we will use evolutionary rate based inference of selection mode, as first introduced by @baker2016 for continuous traits, and later adapted to clock-based rates by @simões2021.

### 1. Import and transform table

```{r,eval=FALSE}
## Import rate table with clade membership (new "clade" column added) 
## from your local directory with "mean" values
rate_table_clades_means3 <- read.csv("RateTable_Means3_Clades.csv", header = TRUE)
```

Below, we use the rate table with clade membership `rate_table_clades_means3` that accompanies `EvoPhylo`.

```{r}
data(rate_table_clades_means3)
```

It is necessary to transform the table from wide to long format with `clock_reshape()`.

```{r}
## Transform table from wide to long format
rates_by_clade <- clock_reshape(rate_table_clades_means3)
```

### 2. Import combined log file from all runs. 

This is produced by using `combine_log()`. Alternatively, users can also use **LogCombiner** from the [BEAST2](https://www.beast2.org/beagle-beast-2-in-cluster/index.html) software package. The first argument passed to `combine_log()` should be a path to the folder containing the log files to be imported and combined.

```{r,eval=FALSE}
## Import all log (.p) files from all runs and combine them, with burn-in = 25% 
## and downsampling to 2.5k trees in each log file
posterior3p <- combine_log("LogFiles3p", burnin = 0.25, downsample = 1000)
```

Below, we use the posterior dataset `posterior3p` that accompanies `EvoPhylo`.

```{r, results='hide'}
data(posterior3p)

## Show first 10 lines of combined log file
head(posterior3p, 10)
```

### 3. Pairwise t-tests of Rate values 

The function `get_pwt_rates()` will produce a table of pairwise t-tests for differences between the mean clockrate value in the posterior and the absolute rate for each tree node. 

```{r, eval = FALSE}
## Get table of pairwise t-tests for difference between the posterior 
## mean and the rate for each tree node
rate_sign_tests <- get_pwt_rates(rate_table_clades_means3, posterior3p)

## Show first 10 lines of table
head(rate_sign_tests, 10)
```
```{r, echo = FALSE}
rate_sign_tests <- get_pwt_rates(rate_table_clades_means3, posterior3p)
t3 <- head(rate_sign_tests, 10)
kableExtra::kbl(t3, caption = "Combined log file") |> 
  kableExtra::kable_styling(font_size = 15, full_width = FALSE,
                            bootstrap_options = "striped", "condensed")
```
```{r, eval=FALSE}
## Export the table
write.csv(rate_sign_tests, file = "RateSign_tests.csv")
```

### 4. Plot selection gradient on the summary tree

Using different thresholds, Identify the selection mode across branches in the tree for each clock partition with `plot_treerates_sgn()`.

```{r, fig.width=8, fig.height=8, fig.align = "center", out.width = "70%"}
## Plot tree using various thresholds for clock partition 1
A1 <- plot_treerates_sgn(
  tree3p, posterior3p, 
  clock = 1,               #Show rates for clock partition 1
  summary = "mean",        #sets summary stats to get from summary tree nodes
  branch_size = 1.5, tip_size = 3,                      #sets size for tree elements
  xlim = c(-450, -260), nbreaks = 8, geo_size = list(3, 3),  #sets limits and breaks for geoscale
  threshold = c("1 SD", "2 SD"))                 #sets threshold for selection mode
A1
```

Plot tree using various thresholds for the other clock partitions and combine them.

```{r, fig.width=20, fig.height=8, fig.align = "default", out.width = "100%"}
## Plot tree using various thresholds for other clock partition and combine them
A2 <- plot_treerates_sgn(
  tree3p, posterior3p, 
  clock = 2,               #Show rates for clock partition 2
  summary = "mean",        #sets summary stats to get from summary tree nodes
  branch_size = 1.5, tip_size = 3,                      #sets size for tree elements
  xlim = c(-450, -260), nbreaks = 8, geo_size = list(3, 3),  #sets limits and breaks for geoscale
  threshold = c("1 SD", "2 SD"))                 #sets threshold for selection mode

A3 <- plot_treerates_sgn(
  tree3p, posterior3p, 
  clock = 3,               #Show rates for clock partition 2
  summary = "mean",        #sets summary stats to get from summary tree nodes
  branch_size = 1.5, tip_size = 3,                      #sets size for tree elements
  xlim = c(-450, -260), nbreaks = 8, geo_size = list(3, 3),  #sets limits and breaks for geoscale
  threshold = c("1 SD", "2 SD"))                 #sets threshold for selection mode

library(patchwork)
A1 + A2 + A3 + plot_layout(nrow = 1)
```
```{eval = FALSE}
## Save your plot to your working directory as a PDF
ggplot2::ggsave("Tree_Sel_3p.pdf", width = 20, height = 8)
```

## References
