---
title: "Character Partitioning"
date: '2022-04-08'
output:
  html_vignette:
    toc: yes
vignette: |
  %\VignetteIndexEntry{Character Partitioning} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---
This vignette explains how to conduct automated morphological character partitioning as a pre-processing step for clock (time-calibrated) Bayesian phylogenetic analysis of morphological data, as introduced by @simões2021.

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, collapse = TRUE)
```

## Installing package **EvoPhylo**

```{r, eval = FALSE}
install.packages("EvoPhylo")
### OR
devtools::install_github("tiago-simoes/EvoPhylo")
```

## Character Partitioning

Load the **EvoPhylo** package

```{r}
library(EvoPhylo)
```


### 1.Generate distance matrix

Generate a Gower distance matrix with `get_gower_dist()` by supplying the file path of a .nex file containing a character data matrix:

```{r, eval = FALSE}
#Load a character data matrix and produce a Gower distance matrix
dist_matrix <- get_gower_dist("DataMatrix.nex", numeric = FALSE)
```

Below, we use the example data matrix `characters` that accompanies `EvoPhylo`.

```{r}
data(characters)

dist_matrix <- get_gower_dist(characters, numeric = FALSE)
```


### 2. Estimate the optimal number of partitions

The optimal number of partitions (clusters) will be first determined using partitioning around medoids (PAM) with Silhouette widths index (Si) using `get_sil_widths()`. The latter will estimate the quality of each PAM cluster proposal relative to other potential clusters.

```{r, fig.width=6, fig.height=4, fig.align = "center", out.width = "70%"}
## Estimate and plot number of cluster against silhouette width
sw <- get_sil_widths(dist_matrix, max.k = 10)
plot(sw, color = "blue", size = 1)

```

Decide on number of clusters based on plot; here, $k = 3$ partitions appears optimal.

### 3. Simple Workflow: 

3.1. Analyze clusters with PAM under chosen $k$ value (from Si) with `make_clusters()`.

3.2. Produce simple cluster graph

3.3. Export clusters/partitions to Nexus file with `cluster_to_nexus()`.

```{r, fig.width=8, fig.height=5, fig.align = "center", out.width = "70%"}
## Generate and vizualize clusters with PAM under chosen k value. 
clusters <- make_clusters(dist_matrix, k = 3)

plot(clusters)
```
```{r, eval = FALSE}
## Write clusters to Nexus file
cluster_to_nexus(clusters, file = "Clusters_Nexus.txt")
```

### 4. Complete Workflow: 

4.1. Analyze clusters with PAM under chosen $k$ value (from Si) with `make_clusters()`.

4.2. Produce a graphic clustering (tSNEs), coloring data points according to PAM clusters, to independently verify PAM clustering. This is set with the `tsne` argument within `make_clusters()`.

4.3. Export clusters/partitions to Nexus file with `cluster_to_nexus()`. This can be copied and pasted into the [Mr. Bayes](https://nbisweden.github.io/MrBayes/) command block.

```{r, fig.width=10, fig.height=7, fig.align = "center", out.width = "100%"}
#User may also generate clusters with PAM and produce a graphic clustering (tSNEs)
clusters <- make_clusters(dist_matrix, k = 3, tsne = TRUE, tsne_dim = 3)

plot(clusters, nrow = 2, max.overlaps = 5)
```
```{r, eval = FALSE}
#Write clusters/partitions in Nexus file format 
cluster_to_nexus(clusters, file = "Clusters_Nexus.txt")
```

## References
