---
title: "Analyzing FBD Parameters"
date: '2022-04-08'
output:
  html_vignette:
    toc: yes
vignette: |
  %\VignetteIndexEntry{Analyzing FBD Parameters} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

This vignette explains how to extract FBD (fossilized birth-death) parameters (*net diversification*, *relative extinction (turnover)*, and *relative fossilization*) estimated from relaxed clock Bayesian inference analyses produced by the program [Mr. Bayes](https://nbisweden.github.io/MrBayes/), as used in @simões2021.

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_knit$set(root.dir = "E:/Git/EvoPhylo/_dev/Examples/")
```


## FBD Parameters Statistics and Plots

Extract evolutionary rate summary statistics from each node from a Bayesian clock (time-calibrate) summary tree produced by Mr. Bayes, store them in a data frame, produce summary tables, and plots:

Open the **EvoPhylo** package

```{r, warning = FALSE}
library(EvoPhylo)
library(openxlsx)
```


### 1. Import combined log file from all runs. 

This is produced by using `import_log`. Alternatively, users can also use **LogCombiner** from the [BEAST2](https://www.beast2.org/beagle-beast-2-in-cluster/index.html) software package. Table is automatically reshaped from wide to long.

```{r}
## Import all log (.p) files from all runs and combine them, with burn-in = 25% and downsampling to 2.5k trees in each log file
posterior3p <- combine_log("examples/MultiClockTree/LogFiles3p", burnin = 0.25, downsample = 1000)

### OR

#OR from example data file
data(posterior3p)

## Reshape imported combined log file from wide to long with FBD_reshape
posterior3p_long<- FBD_reshape(posterior3p)
```


### 2. Summarize FBD parameters by time bin

Summary statistics for each FBD parameter by time bin can be quickly summarized using `FBD_summary`:

```{r}
## Summarize parameters by time bin and analysis
t3.1<- FBD_summary(posterior3p, digits = 2)

kableExtra::kbl(t3.1, caption = "FBD parameters by time bin") |> 
  kableExtra::kable_styling(font_size = 15, full_width = FALSE, bootstrap_options = "striped", "condensed")

## Export the table
write.csv(t3.1, file="FBD_summary.csv")
```


### 3. Plot the distribution of each FBD parameter

Each of (or all) FBD parameter distributions can be plotted by time bin using various plotting alternatives with `FBD_dens_plot`:

```{r, fig.width=8, fig.height=5, fig.align = "center", out.width = "70%"}
## Plot distribution of the desired FBD parameter by time bin with kernel density plot
FBD_dens_plot(posterior3p, parameter = "net_speciation", type = "density", stack = FALSE)
```

```{r, fig.width=8, fig.height=5, fig.align = "center", out.width = "70%"}
## Plot distribution of the desired FBD parameter by time bin with stacked kernel density plot
FBD_dens_plot(posterior3p, parameter = "net_speciation", type = "density", stack = TRUE)
```

```{r, fig.width=4, fig.height=4, fig.align = "center", out.width = "50%"}
## Plot distribution of the desired FBD parameter by time bin with a violin plot
FBD_dens_plot(posterior3p, parameter = "net_speciation", type = "violin", stack = FALSE, color = "red")
```
```{r, fig.width=12, fig.height=4, fig.align = "center", out.width = "100%"}
## Plot distribution of all FBD parameter by time bin with a violin plot
p1<-FBD_dens_plot(posterior3p, parameter = "net_speciation", type = "violin", stack = FALSE, color = "red")
p2<-FBD_dens_plot(posterior3p, parameter = "relative_extinction", type = "violin", stack = FALSE, color = "cyan3")
p3<-FBD_dens_plot(posterior3p, parameter = "relative_fossilization", type = "violin", stack = FALSE, color = "green3")

p1|p2|p3


##(NOT RUN)
## Save your plot to your working directory as a PDF
#ggplot2::ggsave	("Plot_regs.pdf", width=12, height=4)
```


### 4. Test for assumptions 

In this step, users can perform tests for normality and homoscedasticity in data distribution for each of the FBD parameters under consideration. The output will determine whether parametric or nonparametric tests will be performed subsequently.

```{r}
##### Tests for normality and homoscedasticity for each FBD parameter for all time bins

t3.2<- FBD_tests1(posterior3p)

### Export the output table for all tests
openxlsx::write.xlsx(t3.2, file="FBD_Tests1_Assum.xlsx")

### Plot the output table for each test
## Shapiro-Wilk normality test for each parameter 
# Output as separate tables 
kableExtra::kbl(t3.2$shapiro, digits=4, align=c('c','c','c','c'),  caption = "Shapiro-Wilk normality test ") |> 
  kableExtra::kable_styling(font_size = 12, full_width = FALSE, 
                            bootstrap_options = "striped", "condensed")

# OR as single merged table
t3.2$shapiro$net_speciation$bin <- row.names(t3.2$shapiro$net_speciation)  
t3.2$shapiro$relative_extinction$bin <- row.names(t3.2$shapiro$relative_extinction)  
t3.2$shapiro$relative_fossilization$bin <- row.names(t3.2$shapiro$relative_fossilization)  

k1all <- rbind (t3.2$shapiro$net_speciation, t3.2$shapiro$relative_extinction, t3.2$shapiro$relative_fossilization,
                make.row.names = FALSE)

k1<-kableExtra::kbl(k1all, digits=4, caption = "Shapiro-Wilk normality test ") |> 
  kableExtra::kable_styling(font_size = 12, full_width = FALSE, bootstrap_options = "striped", "condensed")
k1

## Bartlett's test for homogeneity of variance 
k2<-kableExtra::kbl(t3.2$bartlett, caption = "Bartlett's test") |> 
  kableExtra::kable_styling(font_size = 12, full_width = FALSE, bootstrap_options = "striped", "condensed")
k2 

## Fligner-Killeen test for homogeneity of variance 
k3<-kableExtra::kbl(t3.2$fligner, caption = "Fligner-Killeen test") |> 
  kableExtra::kable_styling(font_size = 12, full_width = FALSE, bootstrap_options = "striped", "condensed")
k3

```


Deviations from normality can be displayed graphically using `FBD_normality_plot`:

```{r, fig.width=8, fig.height=6, fig.align = "center", out.width = "100%"}
## Visualize deviations from normality and similarity of variances
FBD_normality_plot(posterior3p)

##(NOT RUN)
## Save your plot to your working directory as a PDF
#ggplot2::ggsave	("Plot_normTests.pdf", width=8, height=6)
```


### 5. Test for significant FBD shifts between time bins

Significant shifts in FBD parameters across time bins can be easily tested using parametric (Student's t-test) and nonparametric (Mann-Whitney test) pairwise comparisons with `FBD_tests2`. Both are automatically calculated and the preferred pairwise comparison will be chosen by the user depending on the results of the assumption tests **step #4** (above).

```{r}
##### Test for significant differences between each time bin for each FBD parameter

t3.3<- FBD_tests2(posterior3p)

### Export the output table for all tests
openxlsx::write.xlsx(t3.3, file="FBD_Tests2_Sign.xlsx")

### Plot the output table for each test
## Pairwise t-tests
# Output as separate tables 
kableExtra::kbl(t3.3$t_tests, digits=4, align=c('c','c','c','c'),  caption = "Significant tests ") |> 
  kableExtra::kable_styling(font_size = 10, full_width = FALSE, 
                            bootstrap_options = "striped", "condensed")

# OR as single merged table
k3.3a <- rbind (t3.3$t_tests$net_speciation, t3.3$t_tests$relative_extinction, t3.3$t_tests$relative_fossilization,
                make.row.names = FALSE)

kableExtra::kbl(k3.3a, digits=4, align=c('c','c','c','c'), caption = "Pairwise t-tests") |> 
  kableExtra::kable_styling(font_size = 12, full_width = FALSE, 
                            bootstrap_options = "striped", "condensed")


## Mann-Whitney tests #Use if Tests in step #4 fail assumptions

# Output as separate tables 
kableExtra::kbl(t3.3$mwu_tests, digits=4, align=c('c','c','c','c'), caption = "Mann-Whitney tests") |> 
  kableExtra::kable_styling(font_size = 10, full_width = FALSE, 
                            bootstrap_options = "striped", "condensed")

# OR as single merged table
k3.3b <- rbind (t3.3$mwu_tests$net_speciation, t3.3$mwu_tests$relative_extinction, t3.3$mwu_tests$relative_fossilization,
                make.row.names = FALSE)

kableExtra::kbl(k3.3b, digits=4, align=c('c','c','c','c'), caption = "Mann-Whitney tests") |> 
  kableExtra::kable_styling(font_size = 12, full_width = FALSE, 
                            bootstrap_options = "striped", "condensed")
```
## References
