---
title: "Part 1. Character Partitioning"
output: html_document
---
This vignette explains how to conduct automated morphological character partitioning as a pre-processing step for phylogenetic analysis of morphological data, as introduced by [Simões & Pierce 2021](https://www.nature.com/articles/s41559-021-01532-x). 

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_knit$set(root.dir = "E:/Git/EvoPhylo/_dev/Examples/")
```

# Installing package *EvoPhylo*

```{r, eval = FALSE}
#install.packages("EvoPhylo")
# or
#devtools::install_github("tiago-simoes/EvoPhylo", lib="C:/Program Files/R/R-4.1.0/library")
```

# Character Partitioning

Open the *EvoPhylo* package

```{r}
library(EvoPhylo)
```

Generate distance matrix, analyze clusters with PAM under chosen k value (from Si)
```{r}
#Load character data matrix and produce Gower distance matrix
Dmatrix <- get_gower_dist("DataMatrix.nex", numeric = FALSE)

#Plot number of cluster against silhouette width
sw <- get_sil_widths(Dmatrix, max.k = 10)
plot(sw, color = "red", size=1)
#Decide on number of clusters based on plot; here, k = 3 partitions
```

### Simple Workflow: 
1. Generate distance matrix
2. Analyze clusters with PAM under chosen k value (from Si). 
3. Produce simple cluster graph
4. Exports clusters/partitions to Nexus file

```{r}
#Generate and vizualize clusters with PAM under chosen k value. 
clusters <- make_clusters(Dmatrix, k = 3, tsne = FALSE)
plot(clusters)

#Write clusters to Nexus file
cluster_to_nexus(clusters, file = "Clusters_Nexus.txt")
```

### Complete Workflow: 
1. Generate distance matrix
2. Analyze clusters with PAM under chosen k value (from Si). 
3. Produce a graphic clustering (tSNEs), coloring data points according to PAM clusters, to independently verify PAM clustering.
4. Exports clusters/partitions to Nexus file

```{r}
#User may also generate clusters with PAM and produce a graphic clustering (tSNEs)
clusters <- make_clusters(Dmatrix, k = 3, tsne = TRUE, tsne_dim = 3)
plot(clusters, nrow = 2, max.overlaps=15)

#Write clusters/partitions to Nexus file
cluster_to_nexus(clusters, file = "Clusters_Nexus.txt")
````
