---
title: "FBD parameters"
date: '2022-04-08'
output:
  html_vignette:
    toc: yes
vignette: |
  %\VignetteIndexEntry{Character Partitioning} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---

This vignette explains how to extract FBD (fossilized birth-death) parameters estimated from relaxed clock Bayesian inference analyses produced by the program Mr. Bayes. 

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_knit$set(root.dir = "E:/Git/EvoPhylo/_dev/Examples/")
```
## FBD Parameters Statistics and Plots

Extract evolutionary rate summary statistics from each node from a Bayesian clock (time-calibrate) summary tree produced by Mr. Bayes, store them in a data frame, produce summary tables, and plots:

Open the *EvoPhylo* package

```{r}
library(EvoPhylo)
library(openxlsx)
```

### 1. Import combined log file from all runs. 

This is produced by using `import_log`. Alternatively, users can also use **LogCombiner** from the [BEAST2](https://www.beast2.org/beagle-beast-2-in-cluster/index.html) software package. Table is automatically reshaped from wide to long.

```{r}
#Import all log (.p) files from all runs and combine them, with burn-in = 25% and downsampling to 2.5k trees in each log file
Comb_posterior3p <- import_log("E:/Git/EvoPhylo/_dev/Examples/MultiClockTree/LogFiles3p", burnin = 0.25, downsample = 2500)

#OR

#Import combined log file from all runs (if available)
#Comb_posterior3p <- read.table("E:/Git/EvoPhylo/_dev/Examples/MultiClockTree/3p_CombLog(4runs).p", header = TRUE)

#Reshape Allruns data from wide to long
#FBD_reshape(Comb_posterior3p)
```


### 2. Summarize FBD parameters by time bin
```{r}
#Summarize parameters by time bin and analysis
t3.1<- FBD_summary(Comb_posterior3p, digits = 2)

kableExtra::kbl(t3.1, caption = "FBD parameters by time bin") |> 
  kableExtra::kable_styling(font_size = 15, full_width = FALSE, bootstrap_options = "striped", "condensed")
#Export the table
#write.csv(t3.1, file="FBD_summary.csv")
```


### 3. Plot density of each FBD parameter by time bin

```{r, fig.width=8, fig.height=5, fig.align = "center", out.width = "70%"}
#Plot distribution of the desired FBD parameter by time bin with kernel density plot
FBD_dens_plot(Comb_posterior3p, parameter = "net_speciation", type = "density", stack = FALSE)
```

```{r, fig.width=8, fig.height=5, fig.align = "center", out.width = "70%"}
#Plot distribution of the desired FBD parameter by time bin with stacked kernel density plot
FBD_dens_plot(Comb_posterior3p, parameter = "net_speciation", type = "density", stack = TRUE)
```

```{r, fig.width=4, fig.height=4, fig.align = "center", out.width = "50%"}
#Plot distribution of the desired FBD parameter by time bin with a violin plot
FBD_dens_plot(Comb_posterior3p, parameter = "net_speciation", type = "violin", stack = FALSE, color = "red")
```
```{r, fig.width=12, fig.height=4, fig.align = "center", out.width = "100%"}
#Plot distribution of all FBD parameter by time bin with a violin plot
p1<-FBD_dens_plot(Comb_posterior3p, parameter = "net_speciation", type = "violin", stack = FALSE, color = "red")
p2<-FBD_dens_plot(Comb_posterior3p, parameter = "net_speciation", type = "violin", stack = FALSE, color = "cyan3")
p3<-FBD_dens_plot(Comb_posterior3p, parameter = "net_speciation", type = "violin", stack = FALSE, color = "green3")

p1|p2|p3

#Save your plot to your working directory as a PDF
#ggplot2::ggsave	("Plot_regs.pdf", width=12, height=4)
```


### 4. Test for assumptions 
```{r}
### Tests for normality and homoscedasticity for each FBD parameter for all time bins

## Shapiro-Wilk normality test for each parameter 
# Output as separate tables 
t3.2<- FBD_tests1(Comb_posterior3p)

kableExtra::kbl(t3.2$shapiro, digits=4, align=c('c','c','c','c'),  caption = "Shapiro-Wilk normality test ") |> 
  kableExtra::kable_styling(font_size = 12, full_width = FALSE, 
                            bootstrap_options = "striped", "condensed")

#OR as single merged table

t3.2$shapiro$net_speciation$bin <- row.names(t3.2$shapiro$net_speciation)  
t3.2$shapiro$relative_extinction$bin <- row.names(t3.2$shapiro$relative_extinction)  
t3.2$shapiro$relative_fossilization$bin <- row.names(t3.2$shapiro$relative_fossilization)  

k1all <- rbind (t3.2$shapiro$net_speciation, t3.2$shapiro$relative_extinction, t3.2$shapiro$relative_fossilization,
                make.row.names = FALSE)

k1<-kableExtra::kbl(k1all, digits=4, caption = "Shapiro-Wilk normality test ") |> 
  kableExtra::kable_styling(font_size = 12, full_width = FALSE, bootstrap_options = "striped", "condensed")
k1

#Bartlett's test for homogeneity of variance 
k2<-kableExtra::kbl(t3.2$bartlett, caption = "Bartlett's test") |> 
  kableExtra::kable_styling(font_size = 12, full_width = FALSE, bootstrap_options = "striped", "condensed")
k2 

#Fligner-Killeen test for homogeneity of variance 
k3<-kableExtra::kbl(t3.2$fligner, caption = "Fligner-Killeen test") |> 
  kableExtra::kable_styling(font_size = 12, full_width = FALSE, bootstrap_options = "striped", "condensed")
k3

######

#Export the table
#openxlsx::write.xlsx(t3.2, file="FBD_Tests1_Assum.xlsx")

```


```{r, fig.width=8, fig.height=6, fig.align = "center", out.width = "100%"}
#Visualize deviations from normality and similarity of variances
FBD_normality_plot(Comb_posterior3p)

#Save your plot to your working directory as a PDF
#ggplot2::ggsave	("Plot_normTests.pdf", width=8, height=6)
```


### 5. Test for significant FBD shifts between time bins
```{r}
### Test for significant differences between each time bin for each FBD parameter
#T-tests (output = tables 1-3)
#Mann-Whitney U tests (output = tables 4-6) #Use if Tests in step #4 fail assumptions
FBD_tests2(Comb_posterior3p)

#OR

#Export the table
#openxlsx::write.xlsx(t3.3, file="FBD_Tests2_Sign.xlsx")
```
