---
title: "Character Partitioning"
date: '2022-04-08'
output:
  html_vignette:
    toc: yes
vignette: |
  %\VignetteIndexEntry{Character Partitioning} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
link-citations: true
---
This vignette explains how to conduct automated morphological character partitioning as a pre-processing step for phylogenetic analysis of morphological data, as introduced by [@simões2021].

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_knit$set(root.dir = "E:/Git/EvoPhylo/_dev/Examples/")
```

## Installing package *EvoPhylo*

```{r, eval = FALSE}
#install.packages("EvoPhylo")
# or
#devtools::install_github("tiago-simoes/EvoPhylo", lib="C:/Program Files/R/R-4.1.0/library")
```

## Character Partitioning

Open the *EvoPhylo* package

```{r}
library(EvoPhylo)
```


### 1.Generate distance matrix
```{r}
#Load character data matrix and produce Gower distance matrix
Dmatrix <- get_gower_dist("DataMatrix.nex", numeric = FALSE)
```


### 2. Estimate optimal number of partitions (clusters)
```{r}
#Plot number of cluster against silhouette width
sw <- get_sil_widths(Dmatrix, max.k = 10)
plot(sw, color = "red", size=1)
#Decide on number of clusters based on plot; here, k = 3 partitions
```


### 3. Simple Workflow: 


3.1. Generate distance matrix
3.2. Analyze clusters with PAM under chosen k value (from Si). 
3.3. Produce simple cluster graph
3.4. Exports clusters/partitions to Nexus file


```{r}
#Generate and vizualize clusters with PAM under chosen k value. 
clusters <- make_clusters(Dmatrix, k = 3, tsne = FALSE)
plot(clusters)

#Write clusters to Nexus file
cluster_to_nexus(clusters, file = "Clusters_Nexus.txt")
```


### 4. Complete Workflow: 


4.1. Generate distance matrix
4.2. Analyze clusters with PAM under chosen k value (from Si). 
4.3. Produce a graphic clustering (tSNEs), coloring data points according to PAM clusters, to independently verify PAM clustering.
4.4. Exports clusters/partitions to Nexus file

```{r}
#User may also generate clusters with PAM and produce a graphic clustering (tSNEs)
clusters <- make_clusters(Dmatrix, k = 3, tsne = TRUE, tsne_dim = 3)
plot(clusters, nrow = 2, max.overlaps=15)

#Write clusters/partitions to Nexus file
cluster_to_nexus(clusters, file = "Clusters_Nexus.txt")
````
