---
title: "Evolutionary Rates & Selection Strength"
output: html_document
---

This vignette explains how to extract evolutionary rate parameters estimated from relaxed clock Bayesian inference analyses produced by the program Mr. Bayes. It also shows how to use evolutionary rate based inference of selection strength (or mode) adapted to clock-based rates, as introduced by [Simões & Pierce 2021](https://www.nature.com/articles/s41559-021-01532-x). 

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_knit$set(root.dir = "E:/Git/EvoPhylo/_dev/Examples/")
```

# Extracting Evolutionary Rates Statistics and Plots

Open the *EvoPhylo* package

```{r}
library(EvoPhylo)
```

Extract evolutionary rate summary statistics from each node from a Bayesian clock (time-calibrate) summary tree produced by Mr. Bayes, store them in a data frame, produce summary tables, and plots:

1. First, get rates from tree file and create rate table. 
2. Export the rate table

Workflow:
```{r}
#Import summary tree produced by Mr. Bayes
tree<-treeio::read.mrbayes("Tree_3p(raw).t")

#Get table of clock rates with summary stats for each node in the tree for each relaxed clock partition 
RateTable_Medians_no_clades <- get_clockrate_table(tree, summary = "median")

#Export the rate table
write.csv(RateTable_Medians_no_clades, file="RateTable_Medians.csv")
```

3. Import rate table with customized clade membership added as an extra column
4. Get summary statistics table and plots for each clade by clock 

Workflow:
```{r}
#Import rate table with clade membership 
RateTable_Medians<- read.csv("RateTable_Medians_Clades.csv", header = TRUE)

#Get summary statistics table for each clade by clock 
clockrate_summary(RateTable_Medians, "Sum_RateTable_Medians.csv", digits=2)
```

```{r, echo = FALSE}
clockrate_summary(RateTable_Medians, digits=2) 
```

Plot distributions of rates by clock partition and clade
```{r}
#Overlapping plots
clockrate_dens_plot(RateTable_Medians, stack = FALSE, nrow = 1, scales = "fixed")
```
```{r}
#Stacked plots
clockrate_dens_plot(RateTable_Medians, stack = TRUE, nrow = 1, scales = "fixed")
```

```{r}
#Stacked plots with viridis color scale
clockrate_dens_plot(RateTable_Medians, stack = TRUE, nrow = 1, scales = "fixed")+
    ggplot2::scale_color_viridis_d() +
    ggplot2::scale_fill_viridis_d()
```

Plot linear regressions of rates from two or more clocks

```{r}
#Plot regressions of rates from two clocks
p12 <- clockrate_reg_plot(RateTable_Medians, clock_x = 1, clock_y = 2)
p13 <- clockrate_reg_plot(RateTable_Medians, clock_x = 1, clock_y = 3)
p23 <- clockrate_reg_plot(RateTable_Medians, clock_x = 2, clock_y = 3)

gridExtra::grid.arrange(p12, p13, p23, nrow = 2)

```

# SelectionStrength

1. Import rate table with customized clade membership added as an extra column

Workflow:
```{r}
#Import rate table with clade membership 
RateTable_Medians<- read.csv("RateTable_Medians_Clades.csv", header = TRUE)

RatesByClade <- clock_reshape(RateTable_Medians)

posterior <- read.table("3p_CombLog(4runs)_8k.p", header = TRUE)

#Get matrix of t-tests for difference between the posterior mean and each rate
get_pwt_rates(RatesByClade, posterior$clockrate.all.)

#Plot tree using various thresholds
plot_treerates_sgn(tree, posterior$clockrate.all., clock = 1,
   threshold = c("1 SD", "2 SD", "95%"))

```


#### ST&P_FBD
(Documented)

Workflow:
```{r}
#Reshape Allruns data from wide to long
#import_log()
data("posterior1p")

#Summarize parameters by time bin and analysis
FBD_summary(posterior1p)

#Plot density of parameter by time bin
FBD_dens_plot(posterior1p, parameter = "net_speciation", type = "density")
FBD_dens_plot(posterior1p, parameter = "net_speciation", type = "violin")

#Tests for normality and homoscedasticity for each parameter across time bins
#and analyses
FBD_tests1(posterior1p)

#Visualize deviations from normality and similarity of variances
FBD_normality_plot(posterior1p)

#Test differences in location for each parameter between time bins 
FBD_tests2(posterior1p)

```
