---
title: "Evolutionary Rates & Selection Strength"
output: html_document
---

An R package to implement data pre-processing steps for analyses like those in [Simões & Pierce 2021](https://www.nature.com/articles/s41559-021-01532-x). 

```{r, setup}
knitr::opts_chunk$set(warning = TRUE)
knitr::opts_knit$set(root.dir = "E:/Git/EvoPhylo/_dev/Examples/")
```

# Extracting Evolutionary Rates Statistics and Plots

Extract evolutionary rate summary statistics from each node from a Bayesian clock (time) summary tree produced by Mr. Bayes and stores them in a data frame.

1. First, get rates from tree file and create rate table. 
2. Export the rate table

Workflow:
```{r}
#Import summary tree
tree<-treeio::read.mrbayes("Tree_3p(raw).t")

#Get table of clock rates with summary stats for each node in the tree for each relaxed clock partition 
RateTable_Medians_no_clades <- get_clockrate_table(tree, summary = "median")

#Export the rate table
write.csv(RateTable_Medians_no_clades, file="RateTable_Medians.csv")
```

3. Import rate table with customized clade membership added as an extra column
4. Get summary statistics table and plots for each clade by clock 

Workflow:
```{r}
#Import rate table with clade membership 
RateTable_Medians<- read.csv("RateTable_Medians_Clades.csv", header = TRUE)

#Get summary statistics table for each clade by clock 
clockrate_summary(RateTable_Medians, "Sum_RateTable_Medians.csv", digits=2)
```

```{r, echo = FALSE}
clockrate_summary(RateTable_Medians, digits=2) 
```


```{r}
#Plot distributions of rates by clock and clade
clockrate_dens_plot(RateTable_Medians, stack = TRUE, nrow = 1)

#Plot regressions of rates from two clocks
p12 <- clockrate_reg_plot(RateTable_Medians, clock_x = 1, clock_y = 2)
p13 <- clockrate_reg_plot(RateTable_Medians, clock_x = 1, clock_y = 3)
p23 <- clockrate_reg_plot(RateTable_Medians, clock_x = 2, clock_y = 3)

gridExtra::grid.arrange(p12, p13, p23, nrow = 2)

```

# SelectionStrength


Workflow:
```{r}
data("rate_table_means")
data("tree1p")
RatesByClade <- clock_reshape(RateTable_Medians)
posterior <- read.table("_dev/Examples/SelectionStrength/3p_TipNodeMC_CombLog_8k.p", header = TRUE)

#Get matrix of t-tests for difference between the posterior mean and each rate
get_pwt_rates(RatesByClade, posterior$clockrate.all.)

#Plot tree using various thresholds
plot_treerates_sgn(tree, posterior$clockrate.all., clock = 1,
   threshold = c("1 SD", "2 SD", "95%"))

```


#### ST&P_FBD
(Documented)

Workflow:
```{r}
#Reshape Allruns data from wide to long
#import_log()
data("posterior1p")

#Summarize parameters by time bin and analysis
FBD_summary(posterior1p)

#Plot density of parameter by time bin
FBD_dens_plot(posterior1p, parameter = "net_speciation", type = "density")
FBD_dens_plot(posterior1p, parameter = "net_speciation", type = "violin")

#Tests for normality and homoscedasticity for each parameter across time bins
#and analyses
FBD_tests1(posterior1p)

#Visualize deviations from normality and similarity of variances
FBD_normality_plot(posterior1p)

#Test differences in location for each parameter between time bins 
FBD_tests2(posterior1p)

```
