---
title: "FBD parameters"
output: html_document
---

This vignette explains how to extract FBD (fossilized birth-death) parameters estimated from relaxed clock Bayesian inference analyses produced by the program Mr. Bayes. 

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_knit$set(root.dir = "E:/Git/EvoPhylo/_dev/Examples/")
```

# Extracting Evolutionary Rates Statistics and Plots

Open the *EvoPhylo* package

```{r}
library(EvoPhylo)
```

Extract evolutionary rate summary statistics from each node from a Bayesian clock (time-calibrate) summary tree produced by Mr. Bayes, store them in a data frame, produce summary tables, and plots:

#### FBD Parameters Statistics and Plots
```{r}
#Reshape Allruns data from wide to long
#import_log()
data("posterior1p")

#Summarize parameters by time bin and analysis
FBD_summary(posterior1p)

#Plot density of parameter by time bin
FBD_dens_plot(posterior1p, parameter = "net_speciation", type = "density")
FBD_dens_plot(posterior1p, parameter = "net_speciation", type = "violin")

#Tests for normality and homoscedasticity for each parameter across time bins
#and analyses
FBD_tests1(posterior1p)

#Visualize deviations from normality and similarity of variances
FBD_normality_plot(posterior1p)

#Test differences in location for each parameter between time bins 
FBD_tests2(posterior1p)

```
